Applying Renormalization Group Theory to the Square Well Liquid

### 2014-06-06 papers/thesis-roth/project.tex
figs should work now Tested on elliot. Took 0:18.62.
project report; close to final draft Tested on elliot. Took 0:18.53.

\documentclass[letterpaper,twocolumn,amsmath,amssymb,prb]{revtex4-1}
\usepackage[demo]{graphicx}% Include figure files. When building at home, use the option [demo] to supress actual need of figures
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

% RG stuff
\newcommand{\FbarD}{\ensuremath{\bar{F}_D(T,n,x)}}
\newcommand{\UbarD}{\ensuremath{\bar{U}_D(T,n,x,\lambdaRG)}}

@@

However, understanding liquids has both philosophical and economic
desireability. Liquid water is---as far as we know---essential for
developing and sustaining life. All cellular processes take place in an
aqueous environment. Understanding liquids is in many ways a step
towards understanding the development and existence of life. On the
economic side of the coin, we recognize liquids to be an important
part of a wide array of industrial applications.

@@

  &\equiv F_\text{ref} + a_1(n) + a_2(n)

@@

This is the \emph{high-temperature expansion}, first derived by

@@

The second-order term is quite a bit more complicated. As discussed in
section~\ref{subsec:TPT}, one must have knowledge of higher-order
correlation functions to determine $a_2$. This term describes the
response of the attractive energy due to the compression of the fluid
from the attractive well. Gil-Villegas \emph{et al.} base their
expression on an approximation by Barker and Henderson.\cite{Barker67}
Barker and Henderson's approximation, called the Local Compressibility
Approximation (LCA), considers fluctuations in the number of particles
in the potential well. In this way, the fluctuations in $a_2$ are
related to the pressure and compressibility of the liquid. Given
pressure $P^\text{HS}$ and isothermal compressibility $K^\text{HS} =
kT\left(\partial n /\partial P^\text{HS}\right)_T$, we have:
\begin{align}
  a_2 &= -\pi n \varepsilon^2kT\int_0^\infty r^2\left[\phi(r)\right]^2\frac{\partial n  g^\text{HS}}{\partial P^\text{HS}}(r)\,dr \nonumber \\
  &= -\pi n \varepsilon^2K^\text{HS}\frac{\partial}{\partial n }\left[\int_\sigma^\infty r^2\left[\phi(r)\right]^2 n  g^\text{HS}\,dr\right] \nonumber \\
  &= \frac{1}{2}\varepsilon K^\text{HS}\frac{\partial}{\partial n }\left[-3 n  b^\text{VDW}\varepsilon\int_1^\infty x^2\left[\phi(x)\right]^2 g^\text{HS}(x)\,dx \right]
\end{align}

@@

Then consider a baseline free energy $F_0(T,n)$, which we will take
from TPT. We assume this baseline accounts for all fluctuations of
wavelength $ < \lambda_0$. We then double the size of our phase cell
and add a correction to the free energy $\delta F_1(T,n)$. This
correction accounts for fluctuations with a wavelength up to
$2\lambda_0$. The new free energy, $F_1(T,n)$, is
\begin{align}
  F_1(T,n) &= F_0(T,n) + \delta F_1(T,n)
\end{align}
Then double the length scale again; $\delta F_2(T,n)$ accounts for
fluctuations up to $4\lambda_0$ and
\begin{align}
  F_2(T,n) &= F_1(T,n) + \delta F_2(T,n)
\end{align}
As we continue this process, the $i^{th}$ correction accounts for
fluctuations of $2^i\lambda_0$, and
\begin{align}
  F_i(T,n) &= F_{i-1}(T,n) + \delta F_i(T,n)
\end{align}
In the end, the free energy is
\begin{align}
  F(T,n) = F_0(T,n) + \sum_{i=1}^\infty\delta F_i(T,n) + \Fattr
\end{align}
where $\Fattr$ is the contribution due to the longest-wavelength
density fluctuations. At this stage, I do not entirely understand why
Forte separates out $\Fattr$.

The $\delta F_i(T,n)$ term is a difference in free energies:
\begin{align}
  \delta F_i(T,n) &= -\kT\ln Z - \left( -\kT\ln Z^* \right) \nonumber \\
  &= -\kT\ln\left[ \frac{Z(T,n)}{Z^*(T,n)} \right]
\end{align}
where $Z$ is the partition function for the entire system, and $Z^*$
is the partition function of a reference system consisting only of the
term $\FbarD$. Forte gives these terms as
\begin{multline}
  Z(T,n,x) = \\ \int dx\, \exp\left[-\beta V\left( \FbarD + \UbarD \right)\right] \label{eqn:Z} \\
\end{multline}
\begin{multline}
  Z^*(T,n,x) = \int dx\, \exp\left[ -\beta V\FbarD \right] \label{eqn:Zstar}
\end{multline}
where $x$ is the amplitude of density fluctuations. $\FbarD$ and $\UbarD$ are formulated similar to averages:
\begin{align}
  \FbarD &= \frac{F_{i-1}(T,n+x) + F_{i-1}(T,n-x)}{2} - F_{i-1}(T,n)
\end{align}
\begin{multline}
  \UbarD = \frac{U(T,n+x,\lambdaRG) + U(T,n-x,\lambdaRG)}{2} \\
  - U(T,n,\lambdaRG)
\end{multline}
\FbarD represents the free energy change due to fluctuations if interactions between cells are neglected, and \UbarD is the contrigution of the interactions between cells. Finally, $U(T,n,\lambdaRG)$ is given by Forte as
\begin{multline}
  U(T,n,\lambdaRG) = g_\sigma^\text{HS}(\neff)\alpha(\lambdaRG) \\
  - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha(\lambdaRG)\omega^2(\lambdaRG) \\
  + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha(\lambdaRG)\gamma(\lambdaRG) \label{eqn:URG} % URG! It sounds like it hurt! :P
\end{multline}
with
\begin{align}
  \alpha(\lambdaRG) &= \frac{2}{3}\pi\epsilon\sigma^3(\lambdaRG^3 - 1) \\
  \omega^2(\lambdaRG) &= \frac{1}{5}\sigma^2\frac{\lambdaRG^5 - 1}{\lambdaRG^3 - 1} \\
  \gamma(\lambdaRG) &= \frac{1}{70}\sigma^4\frac{\lambdaRG^7 - 1}{\lambdaRG^3 - 1}
\end{align}
The pair correlation function $g_\sigma(n)$ tells us how likely we
are to find two hard spheres touching, given some density. In
equation~\ref{eqn:URG} we are considering this quantity with an
effective density \neff.

@@

The grand free energy density below the critical temperature has two distinct minima when
plotted versus density. (See fig.~\ref{fig:SW-phi-lowT} for an
example.) The fluid will be in liquid-vapor equilibrium when those
minima have the same grand free energy. I have designed an algorithm
that adjusts \npart\ (that is, adjusts $\mu$) until the system is in
liquid-vapor equilibrium.  The algorithm is as follows; start at some
temperature $T=T_{start}$, with an initial guess for $\npart$. This
initial guess can be read off of a plot, and optimized ``by hand''
with minimal effort.
\begin{enumerate}
  \item Calculate $\nvap$ and $\nliq$ by calving for the minima in $\phi(T_{start},n)$ over appropriate regions
  \item Calculate $\phi(T_{start},\nvap)$ and $\phi(T_{start},\nliq)$ \label{while-start}
  \item Calculate an estimated change in chemical potential using the slope \[\delta\mu = \frac{\phi(T_{start},\nvap) - \phi(T_{start},\nliq)}{\nliq - \nvap}\]
  \item Calculate $\phi^*(T_{start},n) = \phi(T_{start},n) + n\delta\mu$
  \item Calculate a new $\npart$, $\npart^*$, based on $\phi^*(T_{start},n)$ by finding the local maximum between $\nvap$ and $\nliq$
  \item Re-calculate $\nvap$, $\nliq$, $\phi(T_{start},\nvap)$, and $\phi(T_{start},\nliq)$, using $\npart^*$
  \item Go back to step~\ref{while-start} and repeat until the grand free energy at liquid and vapor densityies are sufficiently similar. I used a dimensionless computation tolerance of 10$^{-5}$.
  \item Increase the temperature to $T = T + \delta T$ and repeat using the final $\npart^*$ as our initial guess for the dividing point between the liquid and vapor free energy minima
\end{enumerate}

>>>>

\documentclass[letterpaper,twocolumn,amsmath,amssymb,prb]{revtex4-1}
\usepackage{graphicx}% Include figure files. When building at home, use the option [demo] to supress actual need of figures
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

@@

% RG stuff
\newcommand{\FbarD}{\ensuremath{\bar{F}(T,n,x)}}
\newcommand{\UbarD}{\ensuremath{\bar{U}(T,n,x,\lambdaRG)}}

@@

However, understanding liquids has both philosophical and economic
desirability. Liquid water is---as far as we know---essential for
developing and sustaining life. All cellular processes take place in an
aqueous environment. Understanding liquids is in many ways a step
towards understanding the development and existence of life. On the
economic side of the coin, we recognize liquids to be an important
part of a wide array of industrial applications.

@@

  &= F_\text{ref} + a_1(n) + \beta a_2(n) + \mathcal{O}(\beta^2) \label{eqn:hi-T-exp}

@@

with
  a_1(n) &= \frac{1}{Z_\xi}\int d\rr^N\, \Wtilde e^{-\beta\Vtilde} \\
  a_2(n) &= -\frac{\sigma_{\xi=0}^2}{2}
Equation~\ref{eqn:hi-T-exp} is the \emph{high-temperature expansion}, first derived by

@@

The second-order term is quite a bit more complicated. As discussed in
section~\ref{subsec:TPT}, one must have knowledge of higher-order
correlation functions to exactly determine $a_2$. This term describes the
response of the attractive energy due to the compression of the fluid
from the attractive well. Gil-Villegas \emph{et al.} base their
expression on an approximation by Barker and Henderson.\cite{Barker67}
Barker and Henderson's approximation, called the Local Compressibility
Approximation (LCA), considers fluctuations in the number of particles
in the potential well. In this way, the fluctuations in $a_2$ are
related to the pressure and compressibility of the liquid. Given
pressure $P^\text{HS}$ and isothermal compressibility $K^\text{HS} =
kT\left(\partial n /\partial P^\text{HS}\right)_T$, we have:
\begin{align}
  a_2 &\approx \frac{1}{2}\varepsilon n K^\text{HS}\frac{da_1}{dn}
  %% a_2 &\approx -\pi n \varepsilon^2kT\int_0^\infty r^2\left[\phi(r)\right]^2\frac{\partial n  g^\text{HS}}{\partial P^\text{HS}}(r)\,dr \nonumber \\
  %% &= -\pi n \varepsilon^2K^\text{HS}\frac{\partial}{\partial n }\left[\int_\sigma^\infty r^2\left[\phi(r)\right]^2 n  g^\text{HS}\,dr\right] \nonumber \\
  %% &= \frac{1}{2}\varepsilon K^\text{HS}\frac{\partial}{\partial n }\left[-3 n  b^\text{VDW}\varepsilon\int_1^\infty x^2\left[\phi(x)\right]^2 g^\text{HS}(x)\,dx \right]
\end{align}

@@

Then consider a baseline free energy $F_0(T,n)$, which we will take
from TPT. We assume this baseline accounts for all fluctuations of
wavelength $ < \lambda_0$. We then double the size of our phase cell
and add a correction to the free energy $\delta F_1(T,n)$. This
correction accounts for fluctuations with a wavelength up to
$2\lambda_0$. The new free energy, $F_1(T,n)$, is
\begin{align}
  F_1(T,n) &= F_0(T,n) + \delta F_1(T,n)
\end{align}
Then double the length scale again; $\delta F_2(T,n)$ accounts for
fluctuations up to $4\lambda_0$ and
\begin{align}
  F_2(T,n) &= F_1(T,n) + \delta F_2(T,n)
\end{align}
As we continue this process, the $i^{th}$ correction accounts for
fluctuations of $2^i\lambda_0$, and
\begin{align}
  F_i(T,n) &= F_{i-1}(T,n) + \delta F_i(T,n)
\end{align}
In the end, the free energy is
\begin{align}
  F(T,n) = F_0(T,n) + \Fattr + \sum_{i=1}^\infty\delta F_i(T,n)
\end{align}
where $\Fattr$ is the contribution due to the longest-wavelength
density fluctuations. At this stage, I do not entirely understand why
Forte separates out $\Fattr$.

The $\delta F_i(T,n)$ term is a difference in free energies:
\begin{align}
  \delta F_i(T,n) &= -\kT\ln Z - \left( -\kT\ln Z^* \right) \nonumber \\
  &= -\kT\ln\left[ \frac{Z(T,n)}{Z^*(T,n)} \right]
\end{align}
where $Z$ is the partition function for the entire system, and $Z^*$
is the partition function of a reference system consisting only of the
term $\FbarD$. Forte gives these terms as
\begin{multline}
  Z(T,n,x) = \\ \int dx\, \exp\left[-\beta V\left( \FbarD + \UbarD \right)\right] \label{eqn:Z} \\
\end{multline}
\begin{multline}
  Z^*(T,n,x) = \int dx\, \exp\left[ -\beta V\FbarD \right] \label{eqn:Zstar}
\end{multline}
where $x$ is the amplitude of density fluctuations. $\FbarD$ and $\UbarD$ are formulated similar to averages:
\begin{align}
  \FbarD &= \frac{F_{i-1}(T,n+x) + F_{i-1}(T,n-x)}{2} - F_{i-1}(T,n)
\end{align}
\begin{multline}
  \UbarD = \frac{U(T,n+x,\lambdaRG) + U(T,n-x,\lambdaRG)}{2} \\
  - U(T,n,\lambdaRG)
\end{multline}
\FbarD represents the free energy change due to fluctuations if interactions between cells are neglected, and \UbarD is the contribution of the interactions between cells. Finally, $U(T,n,\lambdaRG)$ is given by Forte as
\begin{multline}
  U(T,n,\lambdaRG) = g_\sigma^\text{HS}(\neff)\alpha(\lambdaRG) \\
  - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha(\lambdaRG)\omega^2(\lambdaRG) \\
  + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha(\lambdaRG)\gamma(\lambdaRG) \label{eqn:URG} % URG! It sounds like it hurt! :P
\end{multline}
with
\begin{align}
  \alpha(\lambdaRG) &= \frac{2}{3}\pi\epsilon\sigma^3(\lambdaRG^3 - 1) \\
  \omega^2(\lambdaRG) &= \frac{1}{5}\sigma^2\frac{\lambdaRG^5 - 1}{\lambdaRG^3 - 1} \\
  \gamma(\lambdaRG) &= \frac{1}{70}\sigma^4\frac{\lambdaRG^7 - 1}{\lambdaRG^3 - 1}
\end{align}
The pair correlation function $g_\sigma(n)$ tells us how likely we
are to find two hard spheres touching, given some density. In
equation~\ref{eqn:URG} we are considering this quantity with an
effective density \neff.

@@

The grand free energy density below the critical temperature has two distinct minima when
plotted versus density. (See fig.~\ref{fig:SW-phi-lowT} for an
example.) The fluid will be in liquid-vapor equilibrium when those
minima have the same grand free energy. I have designed an algorithm
that adjusts \npart\ (that is, adjusts $\mu$) until the system is in
liquid-vapor equilibrium.  The algorithm is as follows; start at some
temperature $T=T_{start}$, with an initial guess for $\npart$. This
initial guess can be read off of a plot, and optimized ``by hand''
with minimal effort.
\begin{enumerate}
  \item Calculate $\nvap$ and $\nliq$ by calving for the minima in $\phi(T_{start},n)$ over appropriate regions
  \item Calculate $\phi(T_{start},\nvap)$ and $\phi(T_{start},\nliq)$ \label{while-start}
  \item Calculate an estimated change in chemical potential using the slope \[\delta\mu = \frac{\phi(T_{start},\nvap) - \phi(T_{start},\nliq)}{\nliq - \nvap}\]
  \item Calculate $\phi^*(T_{start},n) = \phi(T_{start},n) + n\delta\mu$
  \item Calculate a new $n$ particular value, $\npart^*$, based on $\phi^*(T_{start},n)$ by finding the local maximum between $\nvap$ and $\nliq$
  \item Recalculate the minima in the grand free energy using $\npart^*$ to find an improved estimate for \nvap\ and \nliq
  \item Go back to step~\ref{while-start} and repeat until the grand free energy at liquid and vapor densities are sufficiently similar. I used a dimensionless computation tolerance of 10$^{-5}$.
  \item Increase the temperature to $T = T + \delta T$ and repeat using the final $\npart^*$ as our initial guess for the dividing point between the liquid and vapor free energy minima
\end{enumerate}
